---
title: "Getting Started with CORA"
subtitle: "COrrelation-Redundancy-Aware FDR Adjustment with Genomic Applications"
author: "Joanna Zyprych-Walczak"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with CORA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Introduction

The **CORA** package implements the COrrelation-Redundancy-Aware FDR
adjustment for high-throughput genomic studies. The standard
Benjamini–Hochberg (BH) procedure treats all hypotheses as equally
informative regardless of the correlation structure among test statistics.
In practice, genes within the same biological pathways exhibit strong
co-expression, and BH may therefore report clusters of highly correlated
genes as separate discoveries — inflating the DEG list without
proportionally increasing biological information content.

CORA modifies the BH rejection threshold for gene at rank $i$ by
subtracting the accumulated absolute correlation $c_i$ with all
previously ranked genes:

$$\tilde{p}_{(i)}^{\text{CORA}} = \min_{j \geq i}
  \left\{ \frac{m}{j - c_j} \cdot p_{(j)} \right\},
\quad c_i = \sum_{k=1}^{i-1} |r_{\sigma(k),\, \sigma(i)}|$$

where $\sigma$ is the ranking permutation (ascending p-value order) and
$r_{ij}$ is the absolute Pearson correlation between genes $i$ and $j$.

**Key properties:**

- **Under independence** ($c_i = 0$ for all $i$): CORA $\equiv$ BH
- **Under positive correlation**: CORA is more conservative than BH
- **Subset property**: $\mathcal{R}_\text{CORA} \subseteq \mathcal{R}_\text{BH}$
  — CORA never rejects a gene that BH does not reject
- **Empirically validated**: $\text{FDR}_\text{CORA} \leq \alpha$ in all
  tested scenarios (see Zyprych-Walczak, 2026)

# Installation

```{r install, eval=FALSE}
# Install CORA from GitHub
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("zjanna/CORA")
```

# Quick Start

```{r quickstart}
library(CORA)

set.seed(42)
m   <- 500                    # number of genes
# 50 true DEGs (small p-values) + 450 nulls
rawp <- c(runif(50, 0, 0.005), runif(450, 0, 1))

# Block-correlated structure: first 50 genes form a co-expression module
cor_mat <- diag(m)
cor_mat[1:50, 1:50] <- 0.6
diag(cor_mat) <- 1

# Compare all methods
cora_compare(rawp, cor_mat, alpha = 0.05)
```

Under positive block correlation, CORA rejects fewer genes than BH but
more than BY, while the subset property
$\mathcal{R}_\text{CORA} \subseteq \mathcal{R}_\text{BH}$ is always
maintained.

# Core Functions

## `adjust_CORA()` — CORA-adjusted p-values

The primary function takes a vector of raw p-values and the absolute
correlation matrix and returns CORA-adjusted p-values.

```{r adjust_cora}
adj_bh   <- adjust_BH(rawp)
adj_cora <- adjust_CORA(rawp, cor_mat)

cat("BH   rejections:", sum(adj_bh   < 0.05), "\n")
cat("CORA rejections:", sum(adj_cora < 0.05), "\n")
cat("Reduction:      ", round((1 - sum(adj_cora < 0.05) /
                               sum(adj_bh < 0.05)) * 100, 1), "%\n")
```

CORA-adjusted p-values are always $\geq$ BH-adjusted p-values, confirming
the subset property:

```{r subset_property}
all(adj_cora >= adj_bh - 1e-10)   # TRUE: CORA ⊆ BH
```

## `cora_compare()` — Side-by-side comparison

Applies BH, BY, ABH, TSBH, and CORA to the same p-values and returns
a summary table with the number of rejections and the percentage relative
to BH:

```{r compare}
cora_compare(rawp, cor_mat)
```

If no correlation matrix is supplied, CORA is skipped and only the four
standard BH-family methods are returned:

```{r compare_no_cor}
cora_compare(rawp)   # BH, BY, ABH, TSBH only
```

## `cora_summary()` — Detailed report

Returns a list with:

- `$comparison` — method comparison table
- `$reduction` — % DEG reduction of CORA vs BH
- `$top_genes` — raw and adjusted p-values for the top-ranked genes
- `$overlap` — top-N overlap between CORA and BH gene lists

```{r summary}
# Named p-values for gene-level output
names(rawp) <- paste0("gene", seq_along(rawp))

res <- cora_summary(rawp, cor_mat, alpha = 0.05, top_n = 10)

# Method comparison
res$comparison

# DEG reduction vs BH
cat("DEG reduction:", res$reduction, "%\n")

# Top 10 genes: raw p, BH-adjusted, CORA-adjusted, significance flags
head(res$top_genes, 10)

# Overlap between CORA and BH top-N lists
res$overlap
```

The overlap table confirms that CORA and BH agree on the most strongly
significant genes — CORA's effect is concentrated at the **margin** of
the ranked list, where correlated borderline-significant genes accumulate.

# Worked Example: RNA-seq–like Scenario

This example mimics a typical RNA-seq analysis: a large dataset with
pathway-structured co-expression, where CORA's effect is most pronounced.

```{r rnaseq_example}
set.seed(2024)
m         <- 5000   # genes tested
n_true    <- 400    # true DEGs
n_path    <- 8      # number of co-expression pathways
path_size <- 50     # DEGs per pathway

# Simulate raw p-values
rawp_rna <- c(
  runif(n_true,   0, 0.003),  # true positives: small p-values
  runif(m - n_true, 0, 1)     # nulls
)
names(rawp_rna) <- paste0("gene", seq_len(m))

# Build block-diagonal correlation matrix (pathway structure)
cor_rna <- diag(m)
for (k in seq_len(n_path)) {
  idx <- (k - 1) * path_size + seq_len(path_size)
  cor_rna[idx, idx] <- 0.65   # within-pathway correlation
}
diag(cor_rna) <- 1

# Run comparison
result_rna <- cora_compare(rawp_rna, cor_rna)
print(result_rna)
```

```{r rnaseq_plot, fig.cap="DEG counts by method under pathway-clustered correlation. CORA falls between BY (most conservative) and BH (most liberal). Dashed line: number of true DEGs."}
cols <- c(BH = "steelblue", BY = "steelblue", ABH = "steelblue",
          TSBH = "steelblue", CORA = "firebrick")

barplot(
  result_rna$n_rejected,
  names.arg = result_rna$method,
  col       = cols[result_rna$method],
  main      = "DEG counts by method (RNA-seq-like scenario)",
  ylab      = "Number of rejected hypotheses",
  las       = 1,
  ylim      = c(0, max(result_rna$n_rejected) * 1.15)
)
abline(h = n_true, lty = 2, col = "darkgreen", lwd = 2)
legend("topright", legend = "True DEGs", lty = 2, col = "darkgreen",
       lwd = 2, bty = "n")
```

```{r rnaseq_summary}
res_rna <- cora_summary(rawp_rna, cor_rna, top_n = 20)
cat("DEG reduction vs BH:", res_rna$reduction, "%\n\n")
cat("Top-N overlap between CORA and BH:\n")
print(res_rna$overlap)
```

CORA removes borderline-significant correlated genes while retaining
the strongly significant core — the top-N overlap with BH remains high.

# Using CORA with Real Genomic Data

CORA requires two inputs: a vector of raw p-values and the corresponding
absolute correlation matrix. Below are typical workflows for RNA-seq and
microarray data.

## RNA-seq (DESeq2)

```{r rnaseq_real, eval=FALSE}
library(DESeq2)
library(CORA)

# 1. Run DESeq2
dds <- DESeqDataSet(se, design = ~condition)
dds <- dds[rowSums(counts(dds) >= 10) >= 2, ]   # filter low counts
dds <- DESeq(dds)

# 2. Extract raw p-values (replace NA with 1 for independent filtering)
res  <- results(dds)
rawp <- res$pvalue
rawp[is.na(rawp)] <- 1
names(rawp) <- rownames(res)

# 3. Build correlation matrix from VST-normalised expression
vst_mat    <- assay(vst(dds, blind = FALSE))
cor_matrix <- abs(cor(t(vst_mat)))

# 4. Apply CORA
adj_cora  <- adjust_CORA(rawp, cor_matrix)
DEGs_CORA <- names(which(adj_cora < 0.05))
cat("CORA DEGs:", length(DEGs_CORA), "\n")

# 5. Full comparison with all BH-family methods
cora_compare(rawp, cor_matrix)
```

## Microarray

```{r microarray_real, eval=FALSE}
library(CORA)

# expression_matrix: genes x samples (already normalised, e.g. RMA)
# labels: factor vector of class labels

# 1. Compute p-values (two-sample t-tests)
rawp <- apply(expression_matrix, 1, function(x) {
  t.test(x[labels == levels(labels)[1]],
         x[labels == levels(labels)[2]])$p.value
})

# 2. Build absolute correlation matrix
cor_matrix <- abs(cor(t(expression_matrix)))

# 3. Apply CORA and compare
cora_compare(rawp, cor_matrix)
```

# When Does CORA Make a Difference?

CORA's effect depends on two factors (Proposition 1 in the paper):

1. **Number of true DEGs** — more rejections mean more $c_i$ accumulation
2. **Correlation strength** — stronger within-pathway correlations mean
   larger $c_i$ per gene

```{r sensitivity, fig.cap="CORA DEG reduction relative to BH as a function of correlation strength (rho) and number of true DEGs. The effect is strongest when both factors are large simultaneously."}
set.seed(99)
m <- 1000

grid <- expand.grid(
  rho    = c(0.1, 0.3, 0.5, 0.7, 0.9),
  n_true = c(50, 100, 200, 400)
)

grid$reduction <- mapply(function(rho, nt) {
  rawp_g <- c(runif(nt, 0, 0.003), runif(m - nt, 0, 1))
  cm <- diag(m)
  cm[1:nt, 1:nt] <- rho
  diag(cm) <- 1
  n_bh   <- sum(adjust_BH(rawp_g)          < 0.05)
  n_cora <- sum(adjust_CORA(rawp_g, cm) < 0.05)
  if (n_bh == 0) 0 else round((1 - n_cora / n_bh) * 100, 1)
}, grid$rho, grid$n_true)

# Plot
cols_n <- c("50"  = "#d4e6f1",
            "100" = "#85c1e9",
            "200" = "#2e86c1",
            "400" = "#1a5276")

plot(NULL, xlim = c(0, 1), ylim = c(0, 60),
     xlab = expression(rho ~ "(within-pathway correlation)"),
     ylab = "DEG reduction vs BH (%)",
     main = "CORA effect: correlation strength × number of DEGs")

for (nt in c(50, 100, 200, 400)) {
  sub <- grid[grid$n_true == nt, ]
  lines(sub$rho, sub$reduction, col = cols_n[as.character(nt)],
        lwd = 2, type = "b", pch = 16)
}
legend("topleft",
       legend = paste0("n_true = ", c(50, 100, 200, 400)),
       col    = cols_n,
       lwd = 2, pch = 16, bty = "n")
```

**Practical implication**: CORA has the largest impact on RNA-seq datasets
with many correlated DEGs. In the paper's benchmarks, CORA removed 17–23%
of BH discoveries on RNA-seq (e.g., 617 genes on airway) and only 1–14% on
microarrays (e.g., 5 genes on Lung), consistent with RNA-seq yielding more
DEGs with stronger within-pathway correlations.

# Interpreting CORA Results

## Which genes does CORA remove?

CORA removes genes that are (a) near the BH significance boundary and
(b) highly correlated with more strongly significant genes. These genes
carry **redundant biological information** — they are co-expressed with
already-discovered genes and therefore contribute less marginal evidence
for biological signal.

```{r interpretation}
set.seed(7)
m_ex  <- 300
n_deg <- 60

rawp_ex <- c(runif(n_deg, 0, 0.004), runif(m_ex - n_deg, 0, 1))
names(rawp_ex) <- paste0("gene", seq_len(m_ex))

cm_ex <- diag(m_ex)
cm_ex[1:n_deg, 1:n_deg] <- 0.7
diag(cm_ex) <- 1

adj_bh_ex   <- adjust_BH(rawp_ex)
adj_cora_ex <- adjust_CORA(rawp_ex, cm_ex)

# Genes removed by CORA = in BH but not in CORA
removed <- which(adj_bh_ex < 0.05 & adj_cora_ex >= 0.05)
kept    <- which(adj_cora_ex < 0.05)

cat("BH rejections:   ", sum(adj_bh_ex   < 0.05), "\n")
cat("CORA rejections: ", sum(adj_cora_ex < 0.05), "\n")
cat("Removed by CORA: ", length(removed), "\n\n")

cat("Median raw p-value — removed genes:", round(median(rawp_ex[removed]), 4), "\n")
cat("Median raw p-value — kept genes:   ", round(median(rawp_ex[kept]),    4), "\n")
```

Removed genes have higher p-values (closer to the BH significance
boundary) compared to CORA-accepted genes — confirming that CORA acts
at the **margin of significance**, not on the strongly significant core.

## Top-N overlap with BH

Because CORA only removes marginal genes, it agrees with BH almost
completely on the most significant genes:

```{r overlap_interpretation}
res_ex <- cora_summary(rawp_ex, cm_ex, top_n = 20)
cat("Top-N overlap between CORA and BH:\n")
print(res_ex$overlap)
```

This means that standard workflows focusing on the top 50–100 hits by
p-value are essentially unaffected by CORA. The method is most valuable
when analysing the **entire DEG list** (e.g., for pathway enrichment
analysis) or when selecting a fixed-size biomarker panel where redundant
genes should be excluded.

# Session Information

```{r session}
sessionInfo()
```
