---
title: "Getting Started with CORA"
subtitle: "COrrelation-Redundancy-Aware FDR Adjustment with Genomic Applications"
author: "Joanna Zyprych-Walczak"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    pandoc_args: ["--mathjax"]
vignette: >
  %\VignetteIndexEntry{Getting Started with CORA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Introduction

The **CORA** package implements the COrrelation-Redundancy-Aware FDR
adjustment for high-throughput genomic studies. The standard
Benjamini-Hochberg (BH) procedure treats all hypotheses as equally
informative regardless of the correlation structure among test statistics.
In practice, genes within the same biological pathways exhibit strong
co-expression, and BH may therefore report clusters of highly correlated
genes as separate discoveries, inflating the DEG list without
proportionally increasing biological information content.

CORA modifies the BH rejection threshold for the gene at rank `i` by
subtracting the accumulated absolute correlation `c_i` with all
previously ranked genes:

$$\tilde{p}_{(i)}^{\text{CORA}} = \min_{j \geq i}
  \left\{ \frac{m}{j - c_j} \cdot p_{(j)} \right\},
\quad c_i = \sum_{k=1}^{i-1} |r_{\sigma(k),\, \sigma(i)}|$$

where the genes are ordered by ascending p-value and `r` denotes the
absolute Pearson correlation between two genes.

**Key properties:**

- **Under independence** (`c_i = 0` for all genes): CORA is identical to BH
- **Under positive correlation**: CORA is more conservative than BH
- **Subset property**: the set of CORA rejections is always a subset of
  the BH rejections; CORA never rejects a gene that BH does not reject
- **Empirically validated**: CORA controls FDR (`FDR <= alpha`) in all
  tested scenarios (see Zyprych-Walczak, 2026)

# Installation

```{r install, eval=FALSE}
# Install CORA from GitHub
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("zjanna/CORA")
```

# Quick Start

The package bundles a real ovarian cancer microarray dataset
(Welsh et al. 2001), which we use throughout this vignette. It exposes
two objects: `os_raw` (7129 genes by 36 samples) and `L.ova` (the subtype
class label for each sample, three classes of 12 samples each).

```{r quickstart}
library(CORA)

data(welsh2001_ovarian)
grp <- factor(L.ova)

# Keep the 3000 most variable genes (standard variance filtering)
vars <- apply(os_raw, 1, var)
keep <- order(vars, decreasing = TRUE)[1:3000]
expr <- t(scale(t(os_raw[keep, ])))       # standardize each gene (mean 0, SD 1)
rownames(expr) <- paste0("gene", keep)

# Per-gene p-values: three-class F-test (one-way ANOVA)
pvals <- apply(expr, 1, function(y) summary(aov(y ~ grp))[[1]][["Pr(>F)"]][1])
names(pvals) <- rownames(expr)

# Absolute gene-gene correlation matrix
cor_ov <- abs(cor(t(expr)))

# Compare all five methods
cora_compare(pvals, cor_ov, alpha = 0.05)
```

CORA rejects fewer genes than BH (it removes redundant, correlated
discoveries) but many more than the very conservative BY procedure, while
the subset property (CORA rejections are always a subset of BH rejections)
is maintained. The adaptive procedures ABH and TSBH are more liberal than
BH.

```{r quickstart_plot, fig.cap="Number of differentially expressed genes detected by each method on the ovarian cancer data. CORA (red) falls below BH by removing correlated, redundant genes."}
res_cmp <- cora_compare(pvals, cor_ov)
cols <- c(BH = "steelblue", BY = "steelblue", ABH = "steelblue",
          TSBH = "steelblue", CORA = "firebrick")

barplot(
  res_cmp$n_rejected,
  names.arg = res_cmp$method,
  col       = cols[res_cmp$method],
  main      = "DEG counts by method (ovarian cancer data)",
  ylab      = "Number of rejected genes",
  las       = 1,
  ylim      = c(0, max(res_cmp$n_rejected) * 1.15)
)
```

# Core Functions

All examples below reuse the `pvals` and `cor_ov` objects built in the
Quick Start.

## `adjust_CORA()` — CORA-adjusted p-values

The primary function takes a vector of raw p-values and the absolute
correlation matrix and returns CORA-adjusted p-values.

```{r adjust_cora}
adj_bh   <- adjust_BH(pvals)
adj_cora <- adjust_CORA(pvals, cor_ov)

cat("BH   rejections:", sum(adj_bh   <= 0.05), "\n")
cat("CORA rejections:", sum(adj_cora <= 0.05), "\n")
cat("Reduction:      ", round((1 - sum(adj_cora <= 0.05) /
                               sum(adj_bh <= 0.05)) * 100, 1), "%\n")
```

CORA-adjusted p-values are always >= BH-adjusted p-values, confirming
the subset property:

```{r subset_property}
all(adj_cora >= adj_bh - 1e-10)   # TRUE: CORA rejections are a subset of BH
```

## `cora_compare()` — Side-by-side comparison

Applies BH, BY, ABH, TSBH, and CORA to the same p-values and returns
a summary table with the number of rejections and the percentage relative
to BH:

```{r compare}
cora_compare(pvals, cor_ov)
```

If no correlation matrix is supplied, CORA is skipped and only the four
standard BH-family methods are returned:

```{r compare_no_cor}
cora_compare(pvals)   # BH, BY, ABH, TSBH only
```

## `cora_summary()` — Detailed report

Returns a list with:

- `$comparison` — method comparison table
- `$reduction` — % DEG reduction of CORA vs BH
- `$top_genes` — raw and adjusted p-values for the top-ranked genes
- `$overlap` — top-N overlap between CORA and BH gene lists

```{r summary}
res <- cora_summary(pvals, cor_ov, alpha = 0.05, top_n = 10)

# DEG reduction vs BH
cat("DEG reduction:", res$reduction, "%\n")

# Top 10 genes: raw p, BH-adjusted, CORA-adjusted, significance flags
head(res$top_genes, 10)

# Overlap between CORA and BH top-N lists
res$overlap
```

The overlap table confirms that CORA and BH agree on the most strongly
significant genes; CORA's effect is concentrated at the **margin** of
the ranked list, where correlated borderline-significant genes accumulate.

# Using CORA with Your Own Data

CORA requires two inputs: a vector of raw p-values and the corresponding
absolute correlation matrix. Below are typical workflows for RNA-seq and
microarray data.

## RNA-seq (DESeq2)

```{r rnaseq_real, eval=FALSE}
library(DESeq2)
library(CORA)

# 1. Run DESeq2
dds <- DESeqDataSet(se, design = ~condition)
dds <- dds[rowSums(counts(dds) >= 10) >= 2, ]   # filter low counts
dds <- DESeq(dds)

# 2. Extract raw p-values (replace NA with 1 for independent filtering)
res  <- results(dds)
rawp <- res$pvalue
rawp[is.na(rawp)] <- 1
names(rawp) <- rownames(res)

# 3. Build correlation matrix from VST-normalised expression
vst_mat    <- assay(vst(dds, blind = FALSE))
cor_matrix <- abs(cor(t(vst_mat)))

# 4. Apply CORA
adj_cora  <- adjust_CORA(rawp, cor_matrix)
DEGs_CORA <- names(which(adj_cora <= 0.05))
cat("CORA DEGs:", length(DEGs_CORA), "\n")

# 5. Full comparison with all BH-family methods
cora_compare(rawp, cor_matrix)
```

## Microarray

```{r microarray_real, eval=FALSE}
library(CORA)

# expression_matrix: genes x samples (already normalised, e.g. RMA)
# labels: factor vector of class labels

# 1. Compute p-values (two-sample t-tests)
rawp <- apply(expression_matrix, 1, function(x) {
  t.test(x[labels == levels(labels)[1]],
         x[labels == levels(labels)[2]])$p.value
})

# 2. Build absolute correlation matrix
cor_matrix <- abs(cor(t(expression_matrix)))

# 3. Apply CORA and compare
cora_compare(rawp, cor_matrix)
```

# When Does CORA Make a Difference?

CORA's effect depends on two factors (Observation 2 in the paper):

1. **Number of true DEGs** — more rejections mean more `c_i` accumulation
2. **Correlation strength** — stronger within-pathway correlations mean
   larger `c_i` per gene

**Practical implication**: CORA has the largest impact on datasets with
many correlated DEGs. In the paper's benchmarks, CORA removed 17-23%
of BH discoveries on RNA-seq (e.g., 617 genes on airway) and 1-17% on
microarrays (e.g., 17.3% on this ovarian dataset), consistent with
stronger within-pathway correlations producing larger reductions.

# Interpreting CORA Results

## Which genes does CORA remove?

CORA removes genes that are (a) near the BH significance boundary and
(b) highly correlated with more strongly significant genes. These genes
carry **redundant biological information**: they are co-expressed with
already-discovered genes and therefore contribute less marginal evidence
for biological signal.

```{r interpretation}
# Reusing adj_bh and adj_cora from the Core Functions section
removed <- which(adj_bh <= 0.05 & adj_cora > 0.05)   # in BH, not in CORA
kept    <- which(adj_cora <= 0.05)

cat("BH rejections:   ", sum(adj_bh   <= 0.05), "\n")
cat("CORA rejections: ", sum(adj_cora <= 0.05), "\n")
cat("Removed by CORA: ", length(removed), "\n\n")

cat("Median raw p-value, removed genes:", round(median(pvals[removed]), 4), "\n")
cat("Median raw p-value, kept genes:   ", round(median(pvals[kept]),    5), "\n")
```

Removed genes have higher p-values (closer to the BH significance
boundary) than CORA-accepted genes, confirming that CORA acts
at the **margin of significance**, not on the strongly significant core.

## Top-N overlap with BH

Because CORA only removes marginal genes, it agrees with BH almost
completely on the most significant genes (see the `res$overlap` table
above: 98-100% overlap at the top 50-200 genes).

This means that standard workflows focusing on the top 50-100 hits by
p-value are essentially unaffected by CORA. The method is most valuable
when analysing the **entire DEG list** (e.g., for pathway enrichment
analysis) or when selecting a fixed-size biomarker panel where redundant
genes should be excluded.

# Session Information

```{r session}
sessionInfo()
```
